---
title: "Covid Vignette"
author: "Xavier Genelin"
date: "10/3/2021"
output: 
  github_document:
    toc: True
    html_preview: False
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This document is a vignette that'll show how we can retrieve data from an API and perform exploratory analysis. This will use the [Covid API](https://covid19api.com/). We will build our own functions that will collect data from the API and allow a user to interact with it by choosing different endpoints they can retrieve.

# Packages

To be able to interact with the API and manipulate the data we want to return, the following packages will need to be installed and used:

* `tidyverse`: this package will allow us to manipulate and visualize the data  
* `httr`: gives us a response object from the API url   
* `jsonlite`: allows us to interact with the API  
* `countrycode`: this lets us tie the country name to the continent which is defined by the World Bank Development Indicators. The Republic of Kosovo is the only one that seems to be missing the continent. This is located in Europe and will be added    
* `knitr`: used to create a table for summary statistics  

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(countrycode)
library(knitr)
```

This also uses the `state.abb` and `state.name` from the `state` dataset. 

# Functions

The following functions help get the data from the covid API that allows the users to select the type of data they want to see. 

## `covidSummary`

This first function gets access to the covid summary data and allows the user to choose either a global or country summary. The global summary will return one row with data on NewConfirmed, TotalConfirmed, NewDeaths, TotalDeaths, NewRecovered, and TotalRecovered along with the date that this data is from. The country data will show the same data for `r nrow(covidSummary("country"))` countries as well as their country name, country code, and slug. We're also adding in the continent based on the country. If the user writes in a continent while trying to access the global information, it will still return the global information. For both the global and country summaries, this function will create a new variable that calculates the total percent that have died from covid. 

```{r}
# The user can either choose a global summary or country

covidSummary <- function(type, continent = "all"){
  ### you can either select global or country to see the global summary or 
  # the summary by country
  getAPI <- GET("https://api.covid19api.com/summary")
  dat <- fromJSON(rawToChar(getAPI$content))
  
  # This lets the user type in capital letters for the type and continent without erroring it out
  type <- tolower(type)

  
  if(type == "global"){
    output <- as_tibble(dat$Global)
    
    # Creates a new variable that calculates the percentage of deaths globally
    output <- output %>% mutate(percentDeath = round(TotalDeaths / TotalConfirmed * 100, 2), Date = as.Date(Date))
    
  } else if(type == "country"){
    output <- as_tibble(dat$Countries)
    
    output <- output %>% select("Country", "CountryCode", "Slug", "NewConfirmed", 
                                "TotalConfirmed", "NewDeaths", "TotalDeaths", "NewRecovered", "TotalRecovered", "Date")
    
    output$Continent <- countrycode(output$Country, origin = "country.name", destination = "continent")
    
    output$Continent[is.na(output$Continent)] <- "Europe"
    continents <- unique(output$Continent)
    
    # This removes unnecessary columns and creates a new column that calculates the percentage of deaths by country and groups that percentage in 3 different levels
    output <- output %>% select("Country", "NewConfirmed", "TotalConfirmed", "NewDeaths", "TotalDeaths", "NewRecovered", "TotalRecovered", "Date", "Continent") %>% mutate(percentDeath = round(TotalDeaths / TotalConfirmed * 100, 2), Date = as.Date(Date), deathPercentGroup = ifelse(round(TotalDeaths / TotalConfirmed * 100, 2) >= 5, "High", ifelse(round(TotalDeaths / TotalConfirmed * 100, 2) >= 2.5, "Medium", "Low")))
    
    output$deathPercentGroup <- as.factor(output$deathPercentGroup)
    
    levels(output$deathPercentGroup) <- c("Low", "Medium", "High")
      if(continent != "all"){
        continent <- str_to_title(continent)
        output <- output %>% filter(Continent == continent)
      
        if(nrow(output) == 0){
          # If the user selects a continent that isn't in the data set, it provides a list to choose from
          stop(paste("Please select one of the following continents: ", paste(sort(continents), collapse = ", "), sep = " "))
        }
      }
  }
  


  return(output)
}
```

## `getCountry`

The next function will allow the user to either type out the country name, country code, or slug it will return the slug value. This is necessary because country filters need to be the slug, so the user can choose anything and not have later functions error out. So for example, if someone wanted to get data on the United States, they would either have to type in "United States of America", "united-states", "US", or "us".

```{r}
getCountry <- function(country, IDType){
  # This will return a tibble with the country name, the country name in lowercase
  # with dashes instead of spaces (slug), and the country code of length 2
  # For example, if we want the United States:
  # country: United States of America, slug: united-states, id: US
  getAPI <- GET("https://api.covid19api.com/countries")
  dat <- fromJSON(rawToChar(getAPI$content))
  data <- as_tibble(dat)
  
  # This lets the user type in with capital letters but won't error out the function
  IDtype <- tolower(IDType)
  
  if(IDType == "country"){
    output <- data %>% filter(Country == country) %>% select(Slug)
  } else if(IDType == "id"){
    output <- data %>% filter(ISO2 == toupper(country)) %>% select(Slug)
  } else if(IDType == "slug"){
    output <- data %>% filter(slug == country) %>% select(Slug)
  } else {
    stop("ERROR: Invalid IDType. Should be country, id, or slug")
  }
  
  if(nrow(output) == 0){
    stop("ERROR: Country not found. Check spelling, possible capitalization, or IDType")
  }
  return(output)
}

```

## `getState`

This acts in a similar way to the `getCountry` function. Some of the APIs require us to filter down either by time or province after selecting the province. So if we wanted data on the United States, we may need to filter by one of those two things. If we wanted to see Wisconsin's data, that is easy to put into the url and get the data. But for states like West Virginia where there are more than one word in the state's name, this has to be the two letter abbreviation of the state. This function will let the user choose either option for the state if they need to filter down the United States data. There may be a similar issue with provinces in other countries, but there doesn't seem to be anything in R to account for that at this time. This function defaults to the state name and has a built in error message if the user doesn't spell the state corecctly or uses the right abbrevation.

```{r}
getState <- function(state, type = "name"){
  states <- data.frame(abbreviation = state.abb, name = state.name)
  
  # removes any capitalization that the user may input
  type <- tolower(type)
  
  if(type == "name"){
    # This capitalizes the first letter of each word 
    # Prevents the filter from getting nothing if the user doesn't enter it properly
    state <- str_to_title(state)
    output <- states %>% filter(name == state) %>% select(abbreviation)
  } else{
    # Capitalizes each letter if the user doesn't to prevent it from returning nothing
    state <- toupper(state)
    output <- states %>% filter(abbreviation == state) %>% select(abbreviation)
  }
  
  if(nrow(output) == 0){
    stop("EROR: Make sure the state name is spelled correctly or the proper abbrevation")
  }
  
  return(output)
}
```


## `dayOne`

Difference versus day one and by country live is the start date. The day one starts when the first case enters the country until the most recent data while the live starts in january 2020 (probably the first case globally) until the most recent data.

Quick notes
It seems like they can only filter on the time range or province in the url
The date is still there after filtering on the province (i.e. Wisconsin) so they could still see a certain range, but that will need to be fitlered after getting the data 
They also need to have a status of confirmed, deaths, or recovered

```{r}
dayOne <- function(country, IDType, status = "all", province = "all", provType, city = "all", fromDate = "", toDate = ""){
  # This checks to see if there will be an error with the status later on before manipulating the dataset at all
  status <- tolower(status)
  if(!(status %in% c("all", "confirmed", "deaths", "recovered"))){
    stop("Status has to be one of the following: all, confirmed, deaths, recovered")
  }
  
  main <- "https://api.covid19api.com/dayone/country/"
  country <- getCountry(country, IDType)
  
  # This gets the main part of the url for the country the user wants to see data for
  url <- paste0(main, country)
  
  # This first checks to see if the user wants to see a specific type of status and put that into the url
  if(status != "all"){
    url <- paste0(url, "/status/", status)
  } # if they leave the status as all, they can specify the province, city, or date range
  
  # This will return country level data if possible, might return a 
  if(province == "all"){
    getAPI <- GET(url)
    dat <- fromJSON(rawToChar(getAPI$content))
    output <- as_tibble(dat)
    
  } else {
    province <- getState(province, type = provType)
    url <- paste0(url, "?province=", province)
    getAPI <- GET(url)
    dat <- fromJSON(rawToChar(getAPI$content))
    output <- as_tibble(dat)
  }
  
  if(output == "for performance reasons, please specify a province"){
    stop("Please specify a province for this country")
  }
  
  # if there is a status that will change the data frame 
  # this will remove any unnecessary information from the output
  if(status == "all"){
    output <- output %>% select("Country", "Province", "City", "Confirmed", "Deaths", "Recovered", "Active", "Date") %>% mutate(Date = as.Date(Date), deathPercentage = round(Deaths/Confirmed, 2))
  } else {
    output <- output %>% select("Country", "Province", "City", "Cases", "Status", "Date") %>% mutate(Date = as.Date(Date))
  }
  
  if(city != "all"){
    # Create a vector of all the cities for the selected province in case of a wrong input
    cities <- sort(unique(output$City))
    output <- output %>% filter(City == city)
    # If the city isn't in the data set, the above code will return a data frame with 0 rows
    if(nrow(output) == 0){
      stop(paste("ERROR: Please select one of these cities for the selected province: ", paste(cities, collapse = ", "), sep = " "))
    }
  }
  
  # if the user specifies either the toDate and/or fromDate, this will filter the data set 
  if(toDate != ""){
    check <- nrow(output)
    output <- output %>% filter(Date <= as.Date(toDate))
    
    if(nrow(output) == check | nrow(output) == 0){
      stop("Plesae enter date in year-month-day format. Data wasn't filtered on this date")
    }
    
  }
  
  if(fromDate != ""){
    check <- nrow(output)
    output <- output %>% filter(Date >= as.Date(fromDate))
    
    if(nrow(output) == check | nrow(output) == 0){
      stop("Plesae enter date in year-month-day format. Data wasn't filtered on this date")
    }
  }
  
  return(output)
}
```

## `covid`

Create one final function that uses the other functions above to collect whatever data is needed
```{r}
covid <- function(func, ...){
  func <- tolower(func)
  
  if(func == "summary"){
    output <- covidSummary(...)
  } else if(func == "dayone"){
    output <- dayOne(...)
  } else {
    stop("Please select either summary or dayone")
  }
  return(output)
}
```

# Data Exporation

The first thing we can look at is the global numbers of covid. This table will show the total numbers since the beginning while also having the new confirmed, deaths, and recovered by the most recent date. The percentDeath column was created to show the total death percentage for the globe. 

```{r}
print(covid("summary", "global"))
```


Make a scatter plot of values for the country summary with continents in different colors or totals in bars. Maybe both

```{r}
country <- covid("summary", "country")
```


```{r}
plot1 <- ggplot(data = country, aes(x = Continent, y = NewConfirmed)) + 
  geom_bar(stat = "identity") +
  labs(title = "New Confirmed Cases by Continent")

plot2 <- ggplot(data = country, aes(x = Continent, y = TotalConfirmed)) + 
  geom_bar(stat = "identity") +
  labs(title = "Total Confirmed Cases by Continent")
```

```{r}
par(mfrow = c(2,1))
plot1
plot2
```

Plot of the top 15 countries with the highest death percentage with confirmed cases above 100. The bars have the total confirmed cases to give us an idea of what the percentage of deaths represents.

```{r}
check <- country %>% filter(TotalConfirmed > 100) %>% arrange(desc(percentDeath)) %>% slice(1:15)

ggplot(data = check, aes(x = Country, y = percentDeath)) + 
  geom_bar(stat = "identity", aes(fill = Continent)) +
  labs(title = "Top 15 Countries by Percentage of Deaths", y = "Percent of Deaths" ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  geom_text(aes(label = TotalConfirmed), size = 2.5, vjust = -0.2)
```

We can see there's a decent mix of continents that have the highest death percentages overall, but this may not be a good indicator of how they're doing recently. As we can see with Yemen, they have a large percentage but they also have less than 10,000 confirmed cases. This may be a country with a small population so this is still significant, or they have done a good job of containing the virus and reducing spread. This could be something that can be analyze further if necessary. The top percentages look to be a little under 5$\%$, so we check how each continent compares to one another based on the different categories. 

```{r}
table(country$Continent, country$deathPercentGroup)
```



Next we'll look at a scatter plot of the new confirmed cases and the new confirmed deaths for each country. This will be the most recent day's data. 


```{r}
ggplot(country, aes(x = NewConfirmed, y = NewDeaths)) +
  geom_point(aes(color = Continent)) + 
  labs(title = "New Covid Cases vs New Covid Deaths", x = "New Cases", y = "New Deaths")
```

There seems to be a lot of variation in deaths with high number of cases. We can also see that 3 of the top 4 death numbers are from the Americas, with the top one being a European country. We can use the same dataset to find out what those 4 countries are to do a further analysis.

```{r}
country %>% arrange(desc(NewDeaths)) %>% slice(1:4)
```

Now we'll examine the Russian Federation, the United States, Mexico, and Brazil to see how they have been doing with confirmed cases and deaths since their first confirmed covid case. To get country level data, we can use the `dayone` option from the `covid` function. With the United States being as large as it is, we'd need to filter on one of the 50 states and can't get country level data. So we'll look at just Russia, Brazil, and Mexico. 

```{r}
russia <- covid("dayone", "Russian Federation", "country")
brazil <- covid("dayone", "Brazil", "country")
mexico <- covid("dayone", "Mexico", "country")

# combines the 3 datasets into one to put the next couple plots side by side
boxes <- rbind(russia, brazil, mexico)

ggplot(boxes, aes(x = Country, y = Confirmed)) +
  geom_boxplot() +
  labs(title = "Confirmed Cases by Country", y = "Confirmed Cases")
```

```{r}
ggplot(boxes, aes(x = Country, y = Deaths)) +
  geom_boxplot() +
  labs(title = "Deaths by Country")
```

So we can see that Brazil has had much higher confirmed numbers and also more deaths. Since this is a highly populated country, this would make sense. The interesting aspect of this is that Russia has more median number of confirmed cases compared to Mexico, but Mexico has a higher median of deaths than Russia. 

To dive in a little deeper with the United States we need to choose a state. I've been in Wisconsin during the whole pandemic, so I'm interested to see what our numbers look like. This will bring in data for each of the counties in Wisconsin as well, so we can choose to see it at either a state or county level. 

```{r}
wisco <- covid("dayone", "US", "id", status = "all", "wisconsin", "name")
# fitler on trempealeau county
tremp <- wisco %>% filter(City == "Trempealeau") %>% select(Confirmed, Deaths, Recovered, Date)

# get state level data for confirmed, deaths, recovered
wisco <- wisco %>% group_by(Date) %>% summarize(Confirmed = sum(Confirmed), Deaths = sum(Deaths), Recovered = sum(Recovered))
```

First we'll examine this at a state level and then Trempealeau county in Wisconsin (this is my county).

```{r}
ggplot(wisco, aes(x = Confirmed)) +
  geom_histogram(color = "black", fill = "red", bins = 50) +
  labs(title = "Histogram of Confirmed Cases in Wisconsin")
```

```{r}
ggplot(tremp, aes(x = Confirmed)) +
  geom_histogram(color = "black", fill = "red", bins = 50) +
  labs(title = "Histogram of Confirmed Cases in Trempealeau County")
```

Both of these look to be distributed similarly, although on different scales due to the difference in populations. The cases aren't the entire story of this though. There are also people that recover from this daily and also unfortunately die. We can look at Wisconsin

```{r}
badgers <- gather(wisco, status, value, Confirmed, Deaths, Recovered)

output <- badgers %>% group_by(status) %>% summarize(min = min(value), q1 = quantile(value, 0.25), med = median(value), avg = mean(value), q3 = quantile(value, 0.75), max = max(value))

kable(output)
```

It's interesting to see that the recovered numbers are all 0 for the state level. This could be something specific with the API or something else that is going wrong. We do a similar analysis for country levels as well. Let's take a look at how the numbers look for the place where this all started, China.

```{r}
china <- covid("dayone", "China", "country")

china <- china %>% group_by(Date) %>% summarize(Confirmed = sum(Confirmed), Deaths = sum(Deaths), Recovered = sum(Recovered))

results <- gather(china, status, value, Confirmed, Deaths, Recovered) %>% group_by(status) %>% summarize(min = min(value), q1 = quantile(value, 0.25), med = median(value), avg = mean(value), q3 = quantile(value, 0.75), max = max(value))

kable(results)
```

Unlike when we examined the Wisconsin data set, the China data set does have recovered numbers. 


# Wrap Up

This vignette is just a small part of what someone can do with the Covid API. There are other endpoints within the API that one can go to, and within each of the endpoints there are countless things someone can examine with this. This did some exploratory analysis into summary data and country, state, and county level data. 




