---
title: "Covid Vignette"
author: "Xavier Genelin"
date: "10/2/2021"
output: 
  github_document:
    toc: True
    html_preview: False
---

############## Current idea of what to do with this
Rework the live, total, and oneDay functions to allow the user to select a date range, province, and city/city code. Probably just the city because the code was a little odd. Not sure how it's choosing that. I should be able to print the list of options from the choices by using `toString(sort(unique(data$City)))`. Maybe make a function for that too, idk. 

The current issue is some of the urls don't work in the same order by just adding on things at the end. I'll have to go into the API site to see which one should be used for what the user might filter on. 

After that the visuals/EDA should be it for creating the vignette and then move on to the blog and stuff. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

This document is a vignette that'll show how we can retrieve data from an API and perform exploratory analysis. This will use the [Covid API](https://covid19api.com/). We will build our own functions that will collect data from the API and allow a user to interact with it by choosing different endpoints they can retrieve.

# Packages

To be able to interact with the API and manipulate the data we want to return, the following packages will need to be installed and used:

* `tidyverse`: this package will allow us to manipulate and visualize the data  
* `httr`: gives us a response object from the API url   
* `jsonlite`: allows us to interact with the API  
* `countrycode`: this lets us tie the country name to the continent which is defined by the World Bank Development Indicators. The Republic of Kosovo is the only one that seems to be missing the continent. This is located in Europe and will be added     

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(countrycode)
```

# Functions

The following functions help get the data from the covid API that allows the users to select the type of data they want to see. 

## `covidSummary`

This first function gets access to the covid summary data and allows the user to choose either a global or country summary. The global summary will return one row with data on NewConfirmed, TotalConfirmed, NewDeaths, TotalDeaths, NewRecovered, and TotalRecovered along with the date that this data is from. The country data will show the same data for `r nrow(covidSummary("country"))` countries as well as their country name, country code, and slug. We're also adding in the continent based on the country. If the user writes in a continent while trying to access the global information, it will still return the global information. For both the global and country summaries, this function will create a new variable that calculates the total percent that have died from covid. 

```{r}
# The user can either choose a global summary or country

covidSummary <- function(type, continent = "all"){
  ### you can either select global or country to see the global summary or 
  # the summary by country
  getAPI <- GET("https://api.covid19api.com/summary")
  dat <- fromJSON(rawToChar(getAPI$content))
  
  # This lets the user type in capital letters for the type and continent without erroring it out
  type <- tolower(type)

  
  if(type == "global"){
    output <- as_tibble(dat$Global)
    
    output <- output %>% mutate(percentDeath = round(TotalDeaths / TotalConfirmed * 100, 2), Date = as.Date(Date))
    
  } else if(type == "country"){
    output <- as_tibble(dat$Countries)
    
    output <- output %>% select("Country", "CountryCode", "Slug", "NewConfirmed", 
                                "TotalConfirmed", "NewDeaths", "TotalDeaths", "NewRecovered", "TotalRecovered", "Date")
    
    output$Continent <- countrycode(output$Country, origin = "country.name", destination = "continent")
    
    output$Continent[is.na(output$Continent)] <- "Europe"
    continents <- unique(output$Continent)
    
    output <- output %>% select("Country", "NewConfirmed", "TotalConfirmed", "NewDeaths", "TotalDeaths", "NewRecovered", "TotalRecovered", "Date", "Continent") %>% mutate(percentDeath = round(TotalDeaths / TotalConfirmed * 100, 2), Date = as.Date(Date))
    
      if(continent != "all"){
        continent <- str_to_title(continent)
        output <- output %>% filter(Continent == continent)
      
        if(nrow(output) == 0){
          # If the user selects a continent that isn't in the data set, it provides a list to choose from
          stop(paste("Please select one of the following continents: ", paste(sort(continents), collapse = ", "), sep = " "))
        }
      }
  }
  


  return(output)
}
```

## `getCountry`

The next function will allow the user to either type out the country name, country code, or slug it will return the slug value. This is necessary because country filters need to be the slug, so the user can choose anything and not have later functions error out. So for example, if someone wanted to get data on the United States, they would either have to type in "United States of America", "united-states", "US", or "us".

```{r}
getCountry <- function(country, IDType){
  # This will return a tibble with the country name, the country name in lowercase
  # with dashes instead of spaces (slug), and the country code of length 2
  # For example, if we want the United States:
  # country: United States of America, slug: united-states, id: US
  getAPI <- GET("https://api.covid19api.com/countries")
  dat <- fromJSON(rawToChar(getAPI$content))
  data <- as_tibble(dat)
  
  # This lets the user type in with capital letters but won't error out the function
  IDtype <- tolower(IDType)
  
  if(IDType == "country"){
    output <- data %>% filter(Country == country) %>% select(Slug)
  } else if(IDType == "id"){
    output <- data %>% filter(ISO2 == toupper(country)) %>% select(Slug)
  } else if(IDType == "slug"){
    output <- data %>% filter(slug == country) %>% select(Slug)
  } else {
    stop("ERROR: Invalid IDType. Should be country, id, or slug")
  }
  
  if(nrow(output) == 0){
    stop("ERROR: Country not found. Check spelling, possible capitalization, or IDType")
  }
  return(output)
}

```

## `getState`

This acts in a similar way to the `getCountry` function. Some of the APIs require us to filter down either by time or province after selecting the province. So if we wanted data on the United States, we may need to filter by one of those two things. If we wanted to see Wisconsin's data, that is easy to put into the url and get the data. But for states like West Virginia where there are more than one word in the state's name, this has to be the two letter abbreviation of the state. This function will let the user choose either option for the state if they need to filter down the United States data. There may be a similar issue with provinces in other countries, but there doesn't seem to be anything in R to account for that at this time. This function defaults to the state name and has a built in error message if the user doesn't spell the state corecctly or uses the right abbrevation.

```{r}
getState <- function(state, type = "name"){
  states <- data.frame(abbreviation = state.abb, name = state.name)
  
  # removes any capitalization that the user may input
  type <- tolower(type)
  
  if(type == "name"){
    # This capitalizes the first letter of each word 
    # Prevents the filter from getting nothing if the user doesn't enter it properly
    state <- str_to_title(state)
    output <- states %>% filter(name == state) %>% select(abbreviation)
  } else{
    # Capitalizes each letter if the user doesn't to prevent it from returning nothing
    state <- toupper(state)
    output <- states %>% filter(abbreviation == state) %>% select(abbreviation)
  }
  
  if(nrow(output) == 0){
    stop("EROR: Make sure the state name is spelled correctly or the proper abbrevation")
  }
  
  return(output)
}
```


## `dayOne`

Difference versus day one and by country live is the start date. The day one starts when the first case enters the country until the most recent data while the live starts in january 2020 (probably the first case globally) until the most recent data.

Quick notes
It seems like they can only filter on the time range or province in the url
The date is still there after filtering on the province (i.e. Wisconsin) so they could still see a certain range, but that will need to be fitlered after getting the data 
They also need to have a status of confirmed, deaths, or recovered

```{r}
dayOne <- function(country, IDType, status = "all", province = "all", provType, city = "all", fromDate = "", toDate = ""){
  status <- tolower(status)
  if(!(status %in% c("all", "confirmed", "deaths", "recovered"))){
    stop("Status has to be one of the following: all, confirmed, deaths, recovered")
  }
  
  main <- "https://api.covid19api.com/dayone/country/"
  country <- getCountry(country, IDType)
  
  # This gets the main part of the url for the country the user wants to see data for
  url <- paste0(main, country)
  
  # This first checks to see if the user wants to see a specific type of status and put that into the url
  if(status != "all"){
    url <- paste0(url, "/status/", status)
  } # if they leave the status as all, they can specify the province, city, or date range
  
  # This will return country level data if possible, might return a 
  if(province == "all"){
    getAPI <- GET(url)
    dat <- fromJSON(rawToChar(getAPI$content))
    output <- as_tibble(dat)
    
  } else {
    province <- getState(province, type = provType)
    url <- paste0(url, "?province=", province)
    getAPI <- GET(url)
    dat <- fromJSON(rawToChar(getAPI$content))
    output <- as_tibble(dat)
  }
  
  if(output == "for performance reasons, please specify a province"){
    stop("Please specify a province for this country")
  }
  
  # if there is a status that will change the data frame 
  # this will remove any unnecessary information from the output
  if(status == "all"){
    output <- output %>% select("Country", "Province", "City", "Confirmed", "Deaths", "Recovered", "Active", "Date") %>% mutate(Date = as.Date(Date))
  } else {
    output <- output %>% select("Country", "Province", "City", "Cases", "Status", "Date") %>% mutate(Date = as.Date(Date))
  }
  
  if(city != "all"){
    # Create a vector of all the cities for the selected province in case of a wrong input
    cities <- sort(unique(output$City))
    output <- output %>% filter(City == city)
    # If the city isn't in the data set, the above code will return a data frame with 0 rows
    if(nrow(output) == 0){
      stop(paste("ERROR: Please select one of these cities for the selected province: ", paste(cities, collapse = ", "), sep = " "))
    }
  }
  
  # if the user specifies either the toDate and/or fromDate, this will filter the data set 
  if(toDate != ""){
    check <- nrow(output)
    output <- output %>% filter(Date <= as.Date(toDate))
    
    if(nrow(output) == check | nrow(output) == 0){
      stop("Plesae enter date in year-month-day format. Data wasn't filtered on this date")
    }
    
  }
  
  if(fromDate != ""){
    check <- nrow(output)
    output <- output %>% filter(Date >= as.Date(fromDate))
    
    if(nrow(output) == check | nrow(output) == 0){
      stop("Plesae enter date in year-month-day format. Data wasn't filtered on this date")
    }
  }
  
  return(output)
}
```


Testing to see if this will work as an error for the total function

It works for this one!!!!
```{r}
getAPI <- GET("https://api.covid19api.com/dayone/country/united-states/status/confirmed?province=wv")
dat <- fromJSON(rawToChar(getAPI$content))
test <- as_tibble(dat)
test
```


## `covid`

Create one final function that uses the other functions above to collect whatever data is needed
```{r}
covid <- function(func, ...){
  func <- tolower(func)
  
  if(func == "summary"){
    output <- covidSummary(...)
  } else if(func == "dayone"){
    output <- dayOne(...)
  } else {
    stop("Please select either summary or dayone")
  }
  return(output)
}
```

```{r}
covid("Summary", "Country", continent = "Americas")
```




# Data Exporation

Make a scatter plot of values for the country summary with continents in different colors or totals in bars. Maybe both
```{r}
regionData <- covidSummary("country")
regionData
ggplot(data = regionData, aes(x = NewDeaths, y = NewConfirmed)) + 
  geom_point(aes(color = Continent))
```

```{r}
ggplot(data = regionData, aes(x = Continent, y = NewConfirmed)) + 
  geom_bar(stat = "identity") +
  labs(title = "New Confirmed Cases by Continent")

ggplot(data = regionData, aes(x = Continent, y = TotalConfirmed)) + 
  geom_bar(stat = "identity") +
  labs(title = "Total Confirmed Cases by Continent")
```

Trying something quick. Plot the top 10-15 countires with the highest new confirmed or new deaths or something
```{r}
check <- regionData %>% arrange(desc(NewConfirmed)) %>% slice(1:15)

  ggplot(data = check, aes(x = Country, y = NewConfirmed)) + 
  geom_bar(stat = "identity") +
  labs(title = "Total Confirmed Cases by Continent")
```


